---
title: "02-test-pca-w-unknowns"
output: html_notebook
---

14 December 2022


Testing individual-based analyses (pca) with unknown samples from AK.


The haplotype file for the baseline samples comes from `10-complete-downsamp-self...`


I'll read in the baseline data and the unknown data, then filter both appropriately (missing data), and try reformating the dataframe and then converting it to a genid object for adegenet.



```{r load-libraries}
library(tidyverse)
library(adegenet)
library(ape)
library(pegas)
library(RColorBrewer)
library(stringr)
library(rubias)
library(radiator)
library(DescTools)

```



```{r load-data}
# baseline data - curated, 997 indivs
baseline <- readRDS("../new_baseline_data/processed/sebastes_spp_id_baseline_haplotypes.rds")

# remove the 6 loci that had HWE and other issues
to_remove <- read_csv("../data/loci_to_remove.csv")

baseline90 <- baseline %>%
  anti_join(., to_remove)
```



```{r load-data-unkn}
# unknowns
test1 <- read_csv("sebastes_larvae1_reported_diploid_haplotype.csv") %>%
  select(-X1, -ar)

# test1 <- read_csv("krista_sebastes_reported_diploid_haplotype.csv") %>%
#   select(-X1, -ar)

test.df <- test1 %>%
  select(-read.depth.1, -read.depth.2) %>%
  pivot_longer(4:5, names_to = "haplotype", values_to = "allele") %>%
  #pivot_longer(4:5, names_to = "reads", values_to = "depth") %>%
  #select(-haplotype) %>%
  separate(col = haplotype, sep = "\\.", into = c("x", "gene_copy"), remove = T) %>%
  select(-x) %>%
  unique() %>%
  rename(indiv = indiv.ID, species = group) %>%
  select(species, indiv, locus, gene_copy, allele)

test.df$gene_copy <- as.numeric(test.df$gene_copy)
```


```{r}
# fill in NAs for the loci that are missing from the unknowns df
unkn <- test.df %>%
  select(indiv) %>%
  unlist() %>%
  unname() %>%
  expand.grid(indiv = ., locus = unique(baseline90$locus), gene_copy = 1:2, stringsAsFactors = FALSE) %>%
  as_tibble() %>%
  left_join(., test.df) %>%
  arrange(indiv, locus, gene_copy, allele) %>%
  unique() %>%
  mutate(species = "sebastes_spp")

```
It looks like there might be some mismatches in names between the fasta file and the vcf - specifically with the "Plate" loci from ESTs.

That worked well enough for most loci... at least to test for now.


Take another look at missing data
```{r}
unkn %>%
  filter(is.na(allele)) %>%
  group_by(locus) %>%
  tally() %>%
  arrange(desc(n))
```





```{r}
# unkn_format <- test.df %>%
#   pivot_wider(names_from = gene_copy, values_from = haplo) %>%
#   group_by(indiv.ID, locus) %>%
#   #mutate(allele.balance = ifelse(`1` == `2`, NA, allele.balance)) %>%
#   pivot_longer(cols = 9:10, names_to = "gene_copy", values_to = "allele") %>%
#   select(indiv.ID, locus, gene_copy, allele, depth, allele.balance) %>%
#   #mutate(allele.balance = ifelse(gene_copy == 1, 1, NA)) %>%
#   mutate(species = "sebastes_spp") %>%
#   rename(indiv = indiv.ID)
# 
# unkn_format$gene_copy <- as.numeric(unkn_format$gene_copy)
# 
# unkn_format %>%
#   group_by(indiv, locus) %>%
#   tally() %>%
#   filter(n >2)
```


```{r}
# slim that down for the pca
baseline_for_combo <- baseline90 %>%
  select(-collection, -gtseq_run, -id, -repunit, -sample_type, -allele.balance, -depth) %>%
  select(species, indiv, locus, gene_copy, allele) %>%
  filter(indiv != "R001570")

baseline_for_combo 
```


```{r}
# and combine the baseline with unknowns
merged_df <- bind_rows(baseline_for_combo, unkn)


merged_df %>%
  filter(str_detect(indiv, "s"))
```




### Toss out indivs with missing data at more than 25 loci
Now, toss out any individual with fewer than 65 non-missing loci
```{r toss-missers}
no_hi_missers <- merged_df %>% 
  group_by(indiv) %>%
  filter(sum(!is.na(allele)) >= (65*2))


# samples that were tossed?
merged_df %>% 
  group_by(indiv) %>%
  filter(sum(is.na(allele)) >= (65*2)) %>%
  select(indiv) %>%
  unique()



# distribution of missing data?
merged_df %>%
  group_by(indiv) %>%
  filter(is.na(allele)) %>%
  tally() %>%
  filter(str_detect(indiv, "s"))
```

In the meantime, let's move forward with the analysis.

```{r}
# first make integers of the alleles
alle_idxs <- merged_df %>% 
  dplyr::select(indiv, locus, gene_copy, allele) %>%
  group_by(locus) %>%
  mutate(alleidx = as.integer(factor(allele, levels = unique(allele)))) %>%
  ungroup() %>%
  arrange(indiv, locus, alleidx) # rubias can handle NA's, so no need to change them to 0's

# select just the columns to retain 
#alle_idx2 <- alle_idxs[,-7]
  
# and spread the alleles
two_col <- alle_idxs %>%
  #group_by(indiv, locus) %>%
  unite(loc, locus, gene_copy, sep = ".") %>%
  #ungroup() %>%
  select(-allele) %>%
  pivot_wider(names_from = loc, values_from = alleidx) 
  
```

Add the species info back on
```{r}
spp_indiv <- merged_df %>%
  select(species, indiv) %>%
  unique()

two_col_spp <- two_col %>%
  left_join(., spp_indiv) %>%
  select(species, everything())

```

Look at assignment, quickly:


I'll split the data frame that I created (but it needed to be bunged together for the conversion of alleles to integers).

```{r}
rubias_format <- two_col_spp %>%
  mutate(sample_type = ifelse(str_detect(indiv, "R"), "reference", "mixture")) %>%
  mutate(repunit = ifelse(sample_type == "reference", species, NA)) %>%
  rename(collection = species) %>%
  select(sample_type, repunit, collection, everything())

```



```{r split-frames}
#split up the reference and mixture data frames
sp_mix <- rubias_format %>%
  filter(sample_type == "mixture")

sp_ref <- rubias_format %>%
  filter(sample_type == "reference")
```


```{r run-rubias}
rubias_output <- infer_mixture(reference = sp_ref, mixture = sp_mix, gen_start_col = 5)

```


```{r}
rubias_output$indiv_posteriors %>%
  filter(PofZ > 0.95) %>%
  group_by(collection) %>%
  tally()

```




```{r}
#distribution of z-scores
rubias_output$indiv_posteriors %>%
  filter(PofZ > 0.90) %>%
  ggplot(aes(x = z_score)) +
  geom_histogram()

```



Grab the POP to look them individually in a PCA
```{r}
pop_larvae <- rubias_output$indiv_posteriors %>%
  filter(PofZ > 0.95, 
         collection %in% c("alutus", "polyspinis")) %>%
  select(indiv, collection)

```




### PCA


```{r}
# create vectors of indivs and species
spp_labels <- two_col_spp$species
indivs <- two_col_spp$indiv

```


```{r}
# make factor?
two_col_spp$species <- factor(two_col_spp$species)
```


Make the df match the requirements for tidy_genomic_data
```{r}
long_df <- alle_idxs %>%
  select(-allele, -gene_copy) %>%
  left_join(., spp_indiv) %>%
  select(species, everything()) %>%
  rename(INDIVIDUALS = indiv, STRATA = species, MARKERS = locus, GT = alleidx)

```

Genotypes should be coded with 3 integers for each alleles. 6 integers in total for the genotypes. e.g. 001002 or 111333 (for heterozygote individual). 6 integers WITH separator: e.g. 001/002 or 111/333 (for heterozygote individual). The separator can be any of these: "/", ":", "_", "-", ".", and will be removed.


```{r}
#library("DescTools")

# create 3 digit integers from the genotypes
long_df$GT3 <- Format(long_df$GT, ldigits = 3, digits = 0)

head(long_df)

# NAs hold
# long_df %>%
#   filter(is.na(GT3))

# fix NAs
long_df0s <- long_df %>%
  mutate(GT3 = ifelse(is.na(GT3), "000", GT3))
```

Now combine the GT3 column per indiv/marker:

```{r}
# make the genos characters and then try pasting them as strings
long_df0s$GT3 <- as.character(long_df0s$GT3)

long_df3digit <- long_df0s %>%
  group_by(INDIVIDUALS, MARKERS) %>% 
  arrange(GT3, .by_group = TRUE) %>% 
  summarise(GENOTYPE = toString(GT3))
  
# paste strings together
long_df3digit$GENOTYPE <- gsub(", ","",long_df3digit$GENOTYPE)


# add back on species identity as strata
df_for_conversion <- long_df0s %>% 
  select(-GT, -GT3) %>%
  left_join(., long_df3digit) %>%
  unique() %>%
  rename(GT = GENOTYPE) %>%
  mutate(GT = ifelse(GT == "000000", NA, GT))

df_for_conversion$STRATA <- as.factor(df_for_conversion$STRATA)

# check on NAs here
df_for_conversion %>%
  filter(str_detect(INDIVIDUALS, "s"))
```




```{r convert-df-to-genind}
# use the radiator package for this conversion
genind_df <- write_genind(df_for_conversion)
```



Now that the data is a genind object, go ahead and run the PCA.

Make PCA
```{r dataset-pca}
# Allele presence absence data are extracted and NAs replaced using tab:
datasetX <- tab(genind_df, NA.method="mean") # double check that is this the appropriate method.

# make PCA
dataset_pca1 <- dudi.pca(datasetX, center = TRUE, scannf = FALSE, scale=FALSE, nf = 1000)

# colors
mycol <- colorRampPalette(c("palegreen", "darkolivegreen", "darkslategrey", "darkgreen", "deepskyblue", "aquamarine4", "seagreen4", "cadetblue", "cyan", "darkblue", "midnightblue", "blue", "dodgerblue", "darkcyan", "darkslateblue", "slateblue", "steelblue4", "skyblue", "paleturquoise4", "brown", "royalblue", "purple4", "orange", "darkorange", "darkgoldenrod", "chocolate", "tan4", "saddlebrown", "sienna", "navajowhite4", "darkgray", "black", "pink"))(55)

# plot with factor labels
pdf("pdf_outputs/sebastes_larvae1_baseline_pca.pdf", width = 10, height = 10)
s.class(dataset_pca1$li, fac=pop(genind_df), wt = rep(1, length(pop(genind_df))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(mycol,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()

```



## make PCA with POP larvae

```{r}
pop_for_genind <- pop_larvae %>%
  left_join(., df_for_conversion, by = c("indiv" = "INDIVIDUALS")) %>%
  rename(INDIVIDUALS = indiv) %>%
  select(-STRATA) %>%
  rename(STRATA = collection)

pop_genind <- write_genind(pop_for_genind)

```


Make PCA
```{r pop-pca}
# Allele presence absence data are extracted and NAs replaced using tab:
pop.data <- tab(pop_genind, NA.method="mean") # double check that is this the appropriate method.

# make PCA
pop_pca1 <- dudi.pca(pop.data, center = TRUE, scannf = FALSE, scale=FALSE, nf = 1000)

# colors
mycol <- colorRampPalette(c("coral", "darkolivegreen", "darkslategrey", "darkgreen", "deepskyblue", "aquamarine4", "seagreen4", "cadetblue", "cyan", "darkblue", "midnightblue", "blue", "dodgerblue", "darkcyan", "darkslateblue", "slateblue", "steelblue4", "skyblue", "paleturquoise4", "brown", "royalblue", "purple4", "orange", "darkorange", "darkgoldenrod", "chocolate", "tan4", "saddlebrown", "sienna", "navajowhite4", "darkgray", "black", "pink"))(55)

# plot with factor labels
pdf("pdf_outputs/pop_larvae1_pca.pdf", width = 10, height = 10)
s.class(pop_pca1$li, fac=pop(pop_genind), wt = rep(1, length(pop(pop_genind))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(mycol,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()

```
