---
title: "PCA"
output: html_notebook
---
Here is the original, unfiltered PCA code - useful for displaying outliers!
I have also added code for PCA figures for the manuscript, including the inset with hopkinsi outliers and the carnatus/chrysomelas overlap.


```{r load-libraries}
library(tidyverse)
library(adegenet)
library("ape")
library("pegas")
library(RColorBrewer)
library(wordcloud) # for non-overlapping labels

# two_col2 <- read_csv("csv_outputs/sebastes_spp_baseline_two_col.csv")
```

## Make PCA for hopkinsi outliers

Read in genepop file consisting of a subset of species combined from downsampled dataset and hopkinsi unfiltered data
```{r read-genepop-hopkinsi}
obj_hop <- read.genepop("../new_baseline_data/processed/hopkinsi_outliers_v2.gen", ncode = 3)

# set population as a factor
obj_hop@pop <- as.factor(obj_hop@pop)
```

make a basic PCA following the adegenet tutorial
```{r hopkinsi-pca}
# Allele presence absence data are extracted and NAs replaced using tab:
hopX <- tab(obj_hop, NA.method="mean")

# make PCA
hop_pca1 <- dudi.pca(hopX,scannf=FALSE,scale=FALSE)

# colors
cols2 <- colorRampPalette(c("midnightblue", "turquoise"))(14)

# plot with factor labels
pdf("pdf_outputs/hopkinsi_outliers_pca_v1.pdf", width = 10, height = 10)
s.class(hop_pca1$li, fac=pop(obj_hop), cell = 0,
        xax=1, yax=2, col=transp(cols2,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()

```

## hopkinsi without outliers

Read in updated genepop file
```{r read-clean-hopkinsi}
clean_hop <- read.genepop("../new_baseline_data/processed/hopkinsi_wo_outliers.gen", ncode = 3)

# set population as a factor
clean_hop@pop <- as.factor(clean_hop@pop)
```

Make the basic PCA plots
```{r hopkinsi-clean-pca}
# Allele presence absence data are extracted and NAs replaced using tab:
clhopX <- tab(clean_hop, NA.method="mean")

# make PCA
clhop_pca1 <- dudi.pca(clhopX,scannf=FALSE,scale=FALSE)

# plot with factor labels
pdf("pdf_outputs/hopkinsi_wo_outliers_pca.pdf", width = 10, height = 10)
s.class(clhop_pca1$li, fac=pop(clean_hop), cell = 0,
        xax=1, yax=2, col=transp(funky(14),.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()

# Plot without labels
pdf("pdf_outputs/hopkinsi_wo_outliers_no_labels.pdf", width = 10, height = 10)
s.class(clhop_pca1$li, fac=pop(clean_hop), clabel = FALSE,
        cell = 0, axesel=FALSE, cstar=0, cpoint=1, col=transp(funky(14),.8))
dev.off()

```

## PCA demonstrating no distance between chrysomelas and carnatus

Read in genepop file
```{r genepop-chrysomelas-carnatus}

gby <- read.genepop("../new_baseline_data/processed/carnatus_chrysomelas_downsampled_clean.gen", ncode = 3)

# set population as a factor
gby@pop <- as.factor(gby@pop)

```

Make PCAs with GBY and others...
```{r hopkinsi-pca}
# Allele presence absence data are extracted and NAs replaced using tab:
gbyX <- tab(gby, NA.method="mean")

# make PCA
gby_pca1 <- dudi.pca(gbyX,scannf=FALSE,scale=FALSE)

# plot with factor labels
pdf("pdf_outputs/carnatus_chrysomelas_pca.pdf", width = 10, height = 10)
s.class(gby_pca1$li, fac=pop(gby), cell = 0,
        xax=1, yax=2, col=transp(funky(10),.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()

```

# only gopher/black-and-yellow
Read in genepop file
```{r genepop-chrysomelas-carnatus}

gby1 <- read.genepop("../new_baseline_data/processed/only_carnatus_chrysomelas.gen", ncode = 3)

# set population as a factor
gby1@pop <- as.factor(gby1@pop)

```

Make PCAs with GBY
```{r gby-pca}
# Allele presence absence data are extracted and NAs replaced using tab:
gby1X <- tab(gby1, NA.method="mean")

# make PCA
gby1_pca1 <- dudi.pca(gby1X,scannf=FALSE,scale=FALSE)

# plot with factor labels
pdf("pdf_outputs/only_carnatus_chrysomelas_pca.pdf", width = 6, height = 6)
s.class(gby1_pca1$li, fac=pop(gby1), cell = 0,
        xax=1, yax=2, col=transp(c("dark blue", "dark green"),.8),
        axesel=FALSE, cstar=0, cpoint=4)
dev.off()

```


## atrovirens, carnatus, chrysomelas

genepop file
```{r}
kgb <- read.genepop("../new_baseline_data/processed/atrovirens_carnatus_chrysomelas.gen", ncode = 3)
# set population as a factor
kgb@pop <- as.factor(kgb@pop)

```

Make PCAs with KGB
```{r kgb-pca}
# Allele presence absence data are extracted and NAs replaced using tab:
kgbX <- tab(kgb, NA.method="mean")

# make PCA
kgb_pca1 <- dudi.pca(kgbX,scannf=FALSE,scale=FALSE)

# plot with factor labels
pdf("pdf_outputs/atrovirens_carnatus_chrysomelas_pca.pdf", width = 6, height = 6)
s.class(kgb_pca1$li, fac=pop(kgb), cell = 0,
        xax=1, yax=2, col=transp(c("dark blue", "dark green", "orange"),.8),
        axesel=FALSE, cstar=0, cpoint=4)
dev.off()

```

## Entire downsampled dataset

genepop file
```{r dataset-genepop}
dataset1 <- read.genepop("../new_baseline_data/processed/sebastes_downsampled_zscore_genepop3d.gen", ncode = 3)

# set population as a factor
dataset1@pop <- as.factor(dataset1@pop)

```

Make PCA
```{r dataset-pca}
# Allele presence absence data are extracted and NAs replaced using tab:
datasetX <- tab(dataset1, NA.method="mean")

# make PCA
dataset_pca1 <- dudi.pca(datasetX,scannf=FALSE,scale=FALSE)

# colors
mycol <- colorRampPalette(c("palegreen", "darkolivegreen", "darkslategrey", "darkgreen", "deepskyblue", "aquamarine4", "seagreen4", "cadetblue", "cyan", "darkblue", "midnightblue", "blue", "dodgerblue", "darkcyan", "darkslateblue", "slateblue", "steelblue4", "skyblue", "paleturquoise4", "brown", "royalblue", "purple4", "orange", "darkorange", "darkgoldenrod", "chocolate", "tan4", "saddlebrown", "sienna", "navajowhite4", "darkgray", "black"))(54)

# plot with factor labels
pdf("pdf_outputs/clean_dataset_pca_v3.pdf", width = 10, height = 10)
s.class(dataset_pca1$li, fac=pop(dataset1), wt = rep(1, length(pop(dataset1))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(mycol,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()

```
## PCA analysis with 54 eigenvectors (for Chris)

```{r}
data(dataset1)
x.data <- tab(dataset1, freq=TRUE, NA.method="mean")
pca.data <- dudi.pca(x.data, center=TRUE, scale=FALSE, scannf = FALSE, nf = 54)

```


```{r}
pca.data$eig
```
So there are actually 986 eigenvalues, not just 54 (the number of species).
This is because the eigenvalues represent the amount of genetic diversity - measured by the multivariate method - represented by each principal component. 
The amount of variance in the 986 PC is pretty small though (10e-07).

Make a plot that includes the eigenvalues
```{r}
s.class(pca.data$li, fac=pop(dataset1),
        col=transp(funky(54),.6),
        axesel=FALSE, cstar=0, cpoint=3)
add.scatter.eig(pca.data$eig[1:54],3,1,2, ratio=.3)
```



###

```{r}
# colors
mycol <- colorRampPalette(c("orange", "darkcyan", "blue", "darkblue", "purple", "red"))(54)

# plot with factor labels
pdf("pdf_outputs/clean_dataset_pca_v4.pdf", width = 10, height = 10)
s.class(dataset_pca1$li, fac=pop(dataset1), wt = rep(1, length(pop(dataset1))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(mycol, .9),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()
```


## PCA in sections

It's busy and hard to look at the full dataset in a pca... I'm going to try to break it up into pieces to provide better resolution for the groups that are clustered on top of each other.

read in genepop file
```{r atrovirens-group-genepop}
atrovirens1 <- read.genepop("../new_baseline_data/processed/maliger_etc_downsampled.gen", ncode = 3)

# set population as a factor
atrovirens1@pop <- as.factor(atrovirens1@pop)
```

Make PCA
```{r atrovirens-pca}
# Allele presence absence data are extracted and NAs replaced using tab:
atrovirensX <- tab(atrovirens1, NA.method="mean")

# make PCA
atrovirens_pca1 <- dudi.pca(atrovirensX,scannf=FALSE,scale=FALSE)

# colors
grpcols <- colorRampPalette(c("darkblue", "darkgreen", "orange", "brown", "black"))(5)

# plot with factor labels
pdf("pdf_outputs/carnatus_chrysomelas_grp_pca_v2.pdf", width = 5, height = 6)
s.class(atrovirens_pca1$li, fac=pop(atrovirens1), cell = 0,
        xax=1, yax=2, col=transp(grpcols,.8),
        axesel=FALSE, cstar=0, cpoint=2)
dev.off()

```

## The center subset, including hopkinsi, borealis, ovalis, etc.

read in genepop file
```{r}
middle1 <- read.genepop("../new_baseline_data/processed/middle_grp_subset.gen", ncode = 3)

# set population as a factor
middle1@pop <- as.factor(middle1@pop)

```

Make PCA
```{r}
# Allele presence absence data are extracted and NAs replaced using tab:
middleX <- tab(middle1, NA.method="mean")

# make PCA
middle_pca1 <- dudi.pca(middleX,scannf=FALSE,scale=FALSE)

# plot with factor labels
pdf("pdf_outputs/middle_grp_pca.pdf", width = 6, height = 6)
s.class(middle_pca1$li, fac=pop(middle1), cell = 0,
        xax=1, yax=2, col=transp(funky(7),.8),
        axesel=FALSE, cstar=0, cpoint=2)
dev.off()

```

## 



## Experimenting with using ggplot for my PCAs rather than base R

Do I have to use base R for my figure?
Apparently not! 
Which means I can test out the ggrepel package
```{r}
library(ggrepel)
```

struggggling
```{r}
ggplot(atrovirens_pca1$li) +
  geom_point(aes(x = Axis1, y = Axis2, color = pop(atrovirens1)), size = 2, alpha = 0.8) +
  geom_label(aes(x = Axis1, y = Axis2, label = factor(pop(atrovirens1))), stat = "identity") +
  theme_bw() +
  xlab("PC 1") +
  ylab("PC 2") +
  guides(color = guide_legend(title="Species"))

atrovirens_pca1
pop(atrovirens1)
```





## Original, unfiltered data

Read in genepop format (converted from two-column using the pipeline) as a genind object for use with adegenet.
```{r}
obj1 <- read.genepop("../new_baseline_data/processed/sebastes_updated_genepop.gen", ncode = 3)

```

```{r}
# genepop format removes all numbers from sample names
# we want to put them back.

# grab the individual ids from our 2-column data frame
# ids <- two_col2$indiv
# indNames(obj1) <- ids # name the indNames component of the genind object
# 
# # take a look
# head(indNames(obj1),10) 

# set population as a factor
obj1@pop <- as.factor(obj1@pop)

```

Make a summary barplot:
```{r}
# op <- par(mar = c(10,4,4,2) + 0.1)
# pdf("pdf_outputs/sample_barplot.pdf", height = 8, width = 8)
# barplot(table(pop(obj1)), col=funky(57), las=2,
#         xlab="", ylab="Sample size", cex.axis = 0.5)
# par(op)
# title(xlab="Species", line=5, cex.lab=1.2)
# dev.off()
```

summarize the alleles per locus
```{r}
temp <- summary(obj1)

# barplot(table((obj1), col=funky(90), las=3,
#         xlab="locus", ylab="number of alleles"))

```


make a basic PCA following the adegenet tutorial
```{r}
# Allele presence absence data are extracted and NAs replaced using tab:
X <- tab(obj1, NA.method="mean")

# make PCA
pca1 <- dudi.pca(X,scannf=FALSE,scale=FALSE)
#pca2 <- dudi.pca(X,cent=FALSE,scale=FALSE,scannf=FALSE,nf=3)

# set rownames for plot?
indNames(obj1) <- obj1@pop

# plot PCA with wordcloud-esque labels (hard to read)
# pdf("pdf_outputs/sebastes_pca_labels_v2.pdf", width = 10, height = 10)
# colorplot(pca1$li, pca1$li, transp=TRUE, cex=1, font=2, xlab="PC 1", ylab="PC 2")
# abline(v=0,h=0,col="grey", lty=2)
# textplot(pca1$li[,1], pca1$li[,2], words=pop(obj1), cex=0.5, new=FALSE)
# dev.off()
```

```{r}
# plot with factor labels
pdf("pdf_outputs/sebastes_factor_label_v2.pdf", width = 10, height = 10)
s.class(pca1$li, fac=pop(obj1),
        xax=1, yax=2, col=transp(funky(58),.8),
        axesel=FALSE, cstar=0, cpoint=1)
# textplot(pca1$li[,1], pca1$li[,2], words = fac(pop(obj1)), cex = 0.5, show.lines = TRUE)
dev.off()

# how many samples per species?
summary(factor(obj1@pop))

# pca without labels
pdf("pdf_outputs/sebastes_pca_no_labels_v2.pdf", width = 10, height = 10)
s.class(pca1$li, fac=pop(obj1), clabel = FALSE,
        axesel=FALSE, cstar=1, cpoint=1, col=transp(funky(58),.9))
dev.off()

# another version

s.label(pca1$li, cpoint = 1, boxes = FALSE, add.plot = FALSE)

s.label(pca1$li, ppoints.col = "red", plabels = list(box = list(draw = FALSE), optim = TRUE))

s.label(pca1$li, ppoints.col= transp(funky(58), .7), plabels = list(box = list(draw = FALSE), optim = TRUE))

?s.label()

```



```{r}
loadingplot(pca1$c1^2)
```

## other related analyses
```{r}
library("hierfstat")
```
```{r}
fstat(obj1)
```

Look at pairwise Fst between groups
```{r}
obj1.matFst <- pairwise.fst(obj1,res.type="matrix")
```

And then build a tree from those data
```{r}
obj1.tree <- nj(obj1.matFst)
plot(obj1.tree, type="unr", tip.col=funky(nPop(obj1)), font=2)
annot <- round(obj1.tree$edge.length,2)
edgelabels(annot[annot>0], which(annot>0), frame="n")
add.scale.bar()
```









## Attempt to change a data frame into a genind object

```{r}
# test1 <- two_col2 %>% head %>%
#   select(-repunit, -sample_type) %>%
#   select(1:8) %>%
#   unite(loc1, 3:4, sep = ",") %>%
#   unite(loc2, 4:5, sep = ",") %>%
#   unite(loc3, 5:6, sep = ",")
#   
```



```{r}
# convert individuals to rownames
# ids <- test1$indiv
# test1$collection <- as.factor(test1$collection)
# 
# test2 <- test1 %>%
#   select(-collection, -indiv)
# 
# df2genind(test2, ploidy = 2, sep = ",", ind.names = ids, pop = test1$collection )
```

```{r}
# df <- two_col2 %>%
#   select(-repunit, -sample_type, -indiv, -collection) %>%
#   unite(loc1, 3:4, sep = ",") %>%
#   unite(loc2, 4:5, sep = ",") %>%
#   unite(loc3, 5:6, sep = ",")
# 
# for (i in 1:90) {
#   unite(loc{i}, {i}:{i+1}, sep = ",")
#   
# }
# 
# lst = lapply(seq(ncol(df)), function(i) {
#     unite_(df, paste0('loc', i), paste0(i:(i+1)), sep = ",")
# }) 

  
```