---
title: "R Notebook"
output: html_notebook
---

species group specific PCAs to understand patterns of mis-assignment.

```{r}
library(tidyverse)
library(adegenet)
library("ape")
library("pegas")
library(RColorBrewer)
library(wordcloud) # for non-overlapping labels

```

read in genepop data
```{r}
maliger <- read.genepop("../new_baseline_data/processed/maliger_caurinus_atrovirens.gen", ncode = 3)
```

make a basic PCA following the adegenet tutorial
```{r}
# Allele presence absence data are extracted and NAs replaced using tab:
X <- tab(maliger, NA.method="mean")

# make PCA
mal.pca <- dudi.pca(X,scannf=FALSE,scale=FALSE)

# plot PCA
pdf("pdf_outputs/maliger_pca_v2.pdf", width = 7, height = 7)
colorplot(mal.pca$li, mal.pca$li, transp=TRUE, cex=2, font=5, xlab="PC 1", ylab="PC 2")
abline(v=0,h=0,col="grey", lty=2)
textplot(mal.pca$li[,1], mal.pca$li[,2], words=pop(maliger), cex=.6, new=FALSE)
dev.off()
```

Based on this it looks pretty clear that that maliger sample is mis-labeled.

Can we look at the Fst value between maliger and caurinus?
```{r}
library(pegas)
Fst(as.loci(maliger))

matFst <- pairwise.fst(maliger)

factor(maliger@pop) # atrovirens; caurinus; maliger

matFst


```



## southern california species

```{r data}

socal <- read.genepop("../new_baseline_data/processed/southerncalif_spp.gen", ncode = 3)

```
make a basic PCA following the adegenet tutorial
```{r}
# Allele presence absence data are extracted and NAs replaced using tab:
S <- tab(socal, NA.method="mean")

# make PCA
socal.pca <- dudi.pca(S,scannf=FALSE,scale=FALSE)

# plot PCA
pdf("pdf_outputs/socal_pca_v2.pdf", width = 10, height = 10)
colorplot(socal.pca$li, socal.pca$li, transp=TRUE, cex=2, font=5, xlab="PC 1", ylab="PC 2")
abline(v=0,h=0,col="grey", lty=2)
textplot(socal.pca$li[,1], socal.pca$li[,2], words=pop(socal), cex=.6, new=FALSE)
dev.off()
```
The resolution among the southern CA species is challenging - what is the Fst value between each group?

```{r}
# what are the factors?
factor(socal@pop)

# 9 levels: semicinctus; rosaceus; chlorostictus; ensifer; simulator; helvomaculatus; brevispinis; constellatus; paucispinis

socal.fst <- pairwise.fst(socal)

socal.fst
```


## kgbc and friends

Read in the data to increase resolution for that cluster
```{r}
kgbc <- read.genepop("../new_baseline_data/processed/kgbc_plus.gen", ncode = 3)
```

make a basic PCA following the adegenet tutorial
```{r}
# Allele presence absence data are extracted and NAs replaced using tab:
k <- tab(kgbc, NA.method="mean")

# make PCA
kgbc.pca <- dudi.pca(k,scannf=FALSE,scale=FALSE)

# plot PCA
pdf("pdf_outputs/kgbc_pca.pdf", width = 10, height = 10)
colorplot(kgbc.pca$li, kgbc.pca$li, transp=TRUE, cex=2, font=5, xlab="PC 1", ylab="PC 2")
abline(v=0,h=0,col="grey", lty=2)
textplot(kgbc.pca$li[,1], kgbc.pca$li[,2], words=pop(kgbc), cex=.6, new=FALSE)
dev.off()
```

## New PCA after removing the helvomaculatus, brevispinis, simulator, and revising the identity of the single maliger sample that is clearly misidentified and is actually caurinus

```{r updated-pca}
# Read in the data
new.pca <- read.genepop("../new_baseline_data/processed/sebastes_no_helv_sim_brev.gen", ncode = 3)
```

make a basic PCA following the adegenet tutorial
```{r}
# Allele presence absence data are extracted and NAs replaced using tab:
sebs <- tab(new.pca, NA.method="mean")

# make PCA
sebastes.pca <- dudi.pca(sebs,scannf=FALSE,scale=FALSE)

# plot PCA
#pdf("pdf_outputs/clean_sebastes_pca.pdf", width = 10, height = 10)
colorplot(sebastes.pca$li, sebastes.pca$li, transp=TRUE, cex=2, font=5, xlab="PC 1", ylab="PC 2")
abline(v=0,h=0,col="grey", lty=2)
#dev.off()

# rather than a colorplot, let's do a labeled PCA
?s.class

# version 1
s.class(sebastes.pca$li, fac=pop(new.pca), col=funky(54))

# version 2
pdf("pdf_outputs/clean_sebastes_pca.pdf", width = 10, height = 10)
s.class(sebastes.pca$li, fac=pop(new.pca),
        col=transp(funky(54), 0.7),
        axesel=FALSE, cstar = 0.8, cpoint = 3, clabel = 1)
dev.off()
```
In this version of the PCA, hopkinsi is all stretched out. Are the z-scores for any of those samples weird?

Bingo. Looking at `05-cleaner-baseline.Rmd` there are 5 hopkinsi samples with very low z-scores.

## New PCA Fst summary stats
Maybe to be used for ascertainment bias stuff?

```{r}
library("hierfstat")

fstat(new.pca)
```

```{r}
matFst <- pairwise.fst(new.pca,res.type="matrix")
```
(a bunch of markers with missing data for one or more populations were removed.)

Make a tree?
```{r}
tree <- nj(matFst)

pdf("pdf_outputs/test_tree.pdf", height = 10, width = 10)
plot(tree, type="unr", tip.col=funky(nPop(new.pca)), font=1)
annot <- round(tree$edge.length,2)
edgelabels(annot[annot>0], which(annot>0), frame="n")
add.scale.bar()
dev.off()
```

Make a plot-thing?
```{r}
pdf("pdf_outputs/fst_table_paint.pdf", height = 10, width = 10)
table.paint(matFst, col.labels=1:54)
dev.off()
```

And then the boxplot
```{r}
pdf("pdf_outputs/fst_boxplot.pdf", height = 10, width = 10)
temp <- matFst
diag(temp) <- NA
boxplot(temp, col=funky(nPop(new.pca)), las=3,
        xlab="Species", ylab="Fst")
dev.off()
```
I wonder how much of this is influenced by the number of loci that were removed?

New test PCA
```{r}
# Allele presence absence data are extracted and NAs replaced using tab:
sebs <- tab(new.pca, NA.method="mean")

pca.seb <- dudi.pca(sebs, center=TRUE, scale=FALSE, nf = 20)

# PCs 1 and 2
s.class(pca.seb$li, fac=pop(new.pca),
        xax=1, yax=2, col=transp(funky(54),.7),
        axesel=FALSE, cstar=0, cpoint=3)
add.scatter.eig(pca.seb$eig[1:50],3,2,3, ratio=.3)
```
Check out the first two PCs and then PCs 2 and 3

```{r}
# PCs 2 and 3
s.class(pca.seb$li, fac=pop(new.pca),
        xax=2, yax=3, col=transp(funky(54),.7),
        axesel=FALSE, cstar=0, cpoint=3)
add.scatter.eig(pca.seb$eig[1:50],3,2,3, ratio=.3)
```

