---
title: "weird squarespots"
output: html_notebook
---

Something strange is happening with the squarespot rockfish samples:
First, five samples had very low (< -3) z-scores, despite assigning to hopkinsi, and then the metadata for most of the hopkinsi samples says that they are "ADULT" under REPORTED_LIFE_STAGE, but that they are "Larvae, Transferred by hand during extraction process" under the sample comments.

This makes me wonder if these are actually larvae and actually siblings?

Probably the easiest approach is to remove them from the baseline, assuming that I have enough samples of hopkinsi to be relatively unaffected.


```{r}
library(tidyverse)
```

grab the genotype data for just the hopkinsi samples
```{r}
# genotypes
genos <- readRDS("../new_baseline_data/processed/called_genos_na_explicit.rds")

# list of hopkinsi samples
hops <- read_csv("csv_outputs/hopkinsi_samples.csv")
```

```{r}
genos %>%
  right_join(., hops, by = "NMFS_DNA_ID")
```

Or best to do it in COLONY, assuming these are groups of siblings (larvae).


