---
title: "looking for structure in high heterozygosity species"
output: html_notebook
---

One comment at my committee meeting was that some of the species with higher than expected heterozygosity might have substructure, and we can test for that in our data.

The species that came to the attention of the group were:
1. melanostictus
2. ensifer
3. rosaceus

```{r}
library(tidyverse)
library(rubias)
library(adegenet)
library(feather)


# data

# genotypes
genos <- readRDS("../new_baseline_data/processed/called_genos_na_explicit.rds") %>% 
  filter(!str_detect(NMFS_DNA_ID, "N")) # remove Lorne's ambiguous yelloweye samples

# filtered 2-col data
filt_data <- read_csv("csv_outputs/clean_filtered_997_dataset.csv")

# genepop format with outliers
outliers <- read.genepop("../new_baseline_data/processed/sebastes_melanostictus_ensifer_rosaceus.gen", ncode = 3)

# set population as a factor
outliers@pop <- as.factor(outliers@pop)

```

Make PCA
```{r dataset-pca}
# Allele presence absence data are extracted and NAs replaced using tab:
outliersX <- tab(outliers, NA.method="mean")

# make PCA
outliers_pca1 <- dudi.pca(outliersX,scannf=FALSE,scale=FALSE)

# colors
mycol <- colorRampPalette(c("palegreen", "darkolivegreen", "darkslategrey", "darkgreen", "deepskyblue", "aquamarine4", "seagreen4", "cadetblue", "cyan", "darkblue", "midnightblue", "blue", "dodgerblue", "darkcyan", "darkslateblue", "slateblue", "steelblue4", "skyblue", "paleturquoise4", "brown", "royalblue", "purple4", "orange", "darkorange", "darkgoldenrod", "chocolate", "tan4", "saddlebrown", "sienna", "navajowhite4", "darkgray", "black"))(54)

# plot with factor labels
pdf("pdf_outputs/melanostictus_etc_pca.pdf", width = 10, height = 10)
s.class(outliers_pca1$li, fac=pop(outliers), wt = rep(1, length(pop(outliers))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(mycol,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()

```
Huh. Only rosaceus looks questionable. Probably need to actually do the analysis in STRUCTURE.

Grab genotypes for the outliers in 2-column format for structure:
```{r}
filt_data %>%
  filter(collection %in% c("melanostictus", "rosaceus", "chlorostictus", "ensifer")) %>%
  select(-sample_type, -repunit) %>%
  write_csv("csv_outputs/filtered_data_2col_outliers.csv")

```

## Look at hopkinsi outliers for read depth

Eric/Carlos had the thought that maybe the hopkinsi samples are contaminated.
Let's look at the genotype data to see if there's a signature of unexpected read depths there:

```{r}
# just the samples of interest
hopis <- c("R010376", "R010374", "R010377", "R010375", "R010372", "R010370", "R010371", "R010373")
hopkis <- as.tibble(hopis) %>%
  rename(NMFS_DNA_ID = value)

hopkis %>% 
  left_join(., genos) %>%
  select(NMFS_DNA_ID, gtseq_run, id) %>%
  unique()
```

Some of those allele balances look suspicious.
I'll poke into the unfiltered data to see about additional alleles.

The assumption with contaminated samples is that they will likely have at least two characteristics:
1. intermediate allele balance (0.4-0.8) for the second allele in at least some loci (generating heterozygotes from what may have been uncontaminated homozygotes)
2. high read depth at a third allele in at least some loci (corresponding to an intermediate allele balance at this third-ranked allele)

Here we look at the unfiltered genotype data for each of the eight samples to determine if their outlier z-statistic was likely caused by contamination.

ID: R010374	gtseq_run: 11	s825
```{r}
test <- read_feather("../new_baseline_data/feather_files/gtseq11_sebastes_spp_10292018.feather") 

test %>%
  filter(id == "s825") %>%
  group_by(locus) %>%
  select(locus, depth, rank, allele.balance) %>%
  arrange(locus, rank) %>%
  filter(rank > 2) %>%
  filter(depth > 8) %>%
  filter(allele.balance > 0.4 & allele.balance < 0.9)

```
The locus tag_id_1558 looks sketchy, as does tag_id_1598, 1604, 1613, 1652.
Maybe contamination IS the issue.


Take a look at another one:
s849
```{r}
test %>%
  filter(id == "s849") %>%
  group_by(locus) %>%
  select(locus, depth, rank, allele.balance) %>%
  arrange(locus, rank) %>%
  filter(rank > 2) %>%
  filter(depth > 8) %>%
  filter(allele.balance > 0.4 & allele.balance < 0.9)


```


s861
```{r}
test %>%
  filter(id == "s861") %>%
  group_by(locus) %>%
  select(locus, depth, rank, allele.balance) %>%
  arrange(locus, rank) %>%
  filter(rank > 2) %>%
  filter(depth > 8) %>%
  filter(allele.balance > 0.4 & allele.balance < 0.9)

```

s837
```{r}
test %>%
  filter(id == "s837") %>%
  group_by(locus) %>%
  select(locus, depth, rank, allele.balance) %>%
  arrange(locus, rank) %>%
  filter(rank > 2) %>%
  filter(depth > 8) %>%
  filter(allele.balance > 0.4 & allele.balance < 0.9)

```

s801
```{r}
test %>%
  filter(id == "s801") %>%
  group_by(locus) %>%
  select(locus, depth, rank, allele.balance) %>%
  arrange(locus, rank) %>%
  filter(rank > 2) %>%
  filter(depth > 8) %>%
  filter(allele.balance > 0.4 & allele.balance < 0.9)

```

s777
```{r}
test %>%
  filter(id == "s777") %>%
  group_by(locus) %>%
  select(locus, depth, rank, allele.balance) %>%
  arrange(locus, rank) %>%
  filter(rank > 2) %>%
  filter(depth > 8) %>%
  filter(allele.balance > 0.4 & allele.balance < 0.9)
```

s789
```{r}
test %>%
  filter(id == "s789") %>%
  group_by(locus) %>%
  select(locus, depth, rank, allele.balance) %>%
  arrange(locus, rank) %>%
  filter(rank > 2) %>%
  filter(depth > 8) %>%
  filter(allele.balance > 0.4 & allele.balance < 0.9)
```


s813
```{r}
test %>%
  filter(id == "s813") %>%
  group_by(locus) %>%
  select(locus, depth, rank, allele.balance) %>%
  arrange(locus, rank) %>%
  filter(rank > 2) %>%
  filter(depth > 8) %>%
  filter(allele.balance > 0.4 & allele.balance < 0.9)
```

Yeah, it looks like all 8 of those samples have tell-tale signs of contamination - very high read depths for the 3 and 4 alleles and intermediate allele balance. Many of the #2 ranked alleles also have intermediate allele balance.



## Meta data

Remind myself what the meta data for these fish showed:
```{r}
hopkis %>%
  left_join(., meta)
```
That's right these are the weird "ADULT" larvae.

## Rosaceus

Four of the rosaceus samples also looked odd based on the structure plot - like maybe they are also contaminated, or at least have mixed ancestry.

R016831
R016832
R016833
R016835

```{r}
genos %>%
  filter(NMFS_DNA_ID == "R016831")

```
This one is gtseq run 55, s560

```{r}
rosy <- read_feather("../new_baseline_data/feather_files/gtseq55_sebastes_spp_10292018.feather") 

rosy %>%
  filter(id == "s560") %>%
  group_by(locus) %>%
  select(locus, depth, rank, allele.balance) %>%
  arrange(locus, rank) %>%
  filter(rank > 2) %>%
  filter(depth > 8) %>%
  filter(allele.balance > 0.4 & allele.balance < 0.9)

```
That one doesn't look contaminated. The only locus with a third allele at significant depth is one that we know is problematic and remove.

Take a look at another one:
R016832
```{r}
genos %>% 
  filter(NMFS_DNA_ID == "R016832")
```

gtseq_run 55, s572
```{r}
rosy %>%
  filter(id == "s572") %>%
  group_by(locus) %>%
  select(locus, depth, rank, allele.balance) %>%
  arrange(locus, rank) %>%
  filter(rank > 2) %>%
  filter(depth > 8) %>%
  filter(allele.balance > 0.4 & allele.balance < 0.9)
  
```

None.

Check one more:
R016833
```{r}
genos %>%
  filter(NMFS_DNA_ID == "R016833")
```

gtseq_run 55, s489
```{r}
rosy %>%
  filter(id == "s489") %>%
  group_by(locus) %>%
  select(locus, depth, rank, allele.balance) %>%
  arrange(locus, rank) %>%
  filter(rank > 2) %>%
  filter(depth > 8) %>%
  filter(allele.balance > 0.4 & allele.balance < 0.9)

```
Not there either.

Finally, R016835:
```{r}
genos %>%
  filter(NMFS_DNA_ID == "R016835")
```

gtseq_run55, s513
```{r}
rosy %>%
  filter(id == "s513") %>%
  group_by(locus) %>%
  select(locus, depth, rank, allele.balance) %>%
  arrange(locus, rank) %>%
  filter(rank > 2) %>%
  filter(depth > 8) %>%
  filter(allele.balance > 0.4 & allele.balance < 0.9)
```
Alright, those four samples are not contaminated.


