---
title: "Eigenvalue analysis for Chris"
output: html_notebook
---

In my committee meeting, Chris questioned whether the gopher/black-and-yellow cluster wouldn't separate more solidly by with two of the other principal components in the analysis. This is a fair point, and one that I should demonstrate empirically.

```{r load-libraries}
library(tidyverse)
library(adegenet)
library("ape")
library("pegas")
library(RColorBrewer)

```

## PCA trying to separate chrysomelas and carnatus

Read in genepop file for just gopher and black-and-yellow
```{r genepop-chrysomelas-carnatus}
gby1 <- read.genepop("../new_baseline_data/processed/only_carnatus_chrysomelas.gen", ncode = 3)

# set population as a factor
gby1@pop <- as.factor(gby1@pop)
```

PCA with just GBY using adegenet
```{r hopkinsi-pca}
# Allele presence absence data are extracted and NAs replaced using tab:
gby1X <- tab(gby1, NA.method="mean")

# make PCA
gby1_pca1 <- dudi.pca(gby1X, center = TRUE, scannf = FALSE, scale=FALSE, nf = 100)

# plot with factor labels
#pdf("pdf_outputs/carnatus_chrysomelas_pca.pdf", width = 10, height = 10)
s.class(gby1_pca1$li, fac=pop(gby1), cell = 0,
        xax=1, yax=2, col=transp(c("dark blue", "dark green"),.8),
        axesel=FALSE, cstar=0, cpoint=4)
add.scatter.eig(gby1_pca1$eig[1:63],3,1,2, ratio=.3)
#dev.off()

```

```{r}
gby1_pca1$eig
```
For these two species, there are 63 eigenvalues

Using the adegenet manual to describe the PC variance in terms of percentage of the total variance:
```{r}
eig.perc <- 100*gby1_pca1$eig/sum(gby1_pca1$eig)

sum(eig.perc[1:30])
```

```{r}
pc1 <- gby1_pca1$li[,1]
var(pc1)
```


Let's see if plotting any of the other PCs will separate gopher/black-and-yellow.
```{r}
# PC 2 and 3
s.class(gby1_pca1$li, fac=pop(gby1), cell = 0,
        xax=2, yax=3, col=transp(c("dark blue", "dark green"),.8),
        axesel=FALSE, cstar=0, cpoint=4)
add.scatter.eig(gby1_pca1$eig[1:63],3,2,3, ratio=.3)

```


Here I'll cycle over the sets of sequential PCs, e.g. PC1 and PC2, PC2 and PC3, etc.
```{r}
pdf("pdf_outputs/carnatus_chrysomelas_pca_tile.pdf", width = 10, height = 10)

par(mfrow = c(3,3))
# name a vector for the first 20 PCs
a <- 1:10
for (i in seq_along(a)) {
s.class(gby1_pca1$li, fac=pop(gby1), cell = 0,
        xax=a[i], yax=a[i+1], col=transp(c("dark blue", "dark green"),.8),
        axesel=FALSE, cstar=0, cpoint=4)
add.scatter.eig(gby1_pca1$eig[1:63],3,a[i],a[i+1], ratio=.3)
}
dev.off()
```

What about comparing PC1 with each other PC (for the first 20)?
```{r pc1}
# using PC1 as the x axis in each case
par(mfrow = c(3,3))
for (i in seq_along(a)) {
s.class(gby1_pca1$li, fac=pop(gby1), cell = 0,
        xax=1, yax=a[i+1], col=transp(c("dark blue", "dark green"),.8),
        axesel=FALSE, cstar=0, cpoint=4)
add.scatter.eig(gby1_pca1$eig[1:63],3,1,a[i+1], ratio=.3)
}
```

# Using PC 2 as the x-axis in each plot
```{r pc2}
par(mfrow = c(3,3))
a <- 3:11
for (i in seq_along(a)) {
s.class(gby1_pca1$li, fac=pop(gby1), cell = 0,
        xax=2, yax=a[i], col=transp(c("dark blue", "dark green"),.8),
        axesel=FALSE, cstar=0, cpoint=4)
add.scatter.eig(gby1_pca1$eig[1:63],3,1,a[i+1], ratio=.3)
}


```

## Mean of the values of each PC for gopher/black-and-yellow

Another way to look at these data is by calculating the mean of the values of each PC for all the gopher individuals and all of the black-and-yellow individuals.

And then plot those means on the y-axis, with the x-axis showing which principal component it is.  I suspect that this will show some separation in the means from the first principal component, and very little after that for any of the others.

```{r species-order}
gby1$pop
```
There we can see that samples 1-32 are chrysomelas and 33-64 are carnatus.

What I want is to take the mean of each column for the first 32 samples and then the mean of each column for the 33-64 samples.

```{r}
# turn df into a tibble,
# sort by the species groups, and then 
# calculate the mean
gby_pc_means <- rownames_to_column(gby1_pca1$li, var = "id") %>%
  mutate(species = ifelse(id < 33, "chrysomelas", "carnatus")) %>%
  group_by(species) %>%
  summarise_all(funs(mean))

gby_pc_means
```

So then I want to plot those means on the y-axis and have PC on the x-axis.

```{r}
mean_df <- gby_pc_means %>%
  gather(pc, mean, 3:65) %>%
  select(-id) 

# rename the Axis to just the PC number
mean_df$pc <- str_replace(mean_df$pc, "Axis", "")
mean_df$pc <- as.numeric(mean_df$pc)
```



```{r gby-means-plot}
mean_plot <- mean_df %>%
  ggplot() +
  geom_point(aes(x = pc, y = mean, color = species)) +
  theme_bw() +
  scale_color_manual(values = c("dark green", "dark blue"))

mean_plot +
  scale_x_continuous(name = "principal component", breaks=seq(0,65,10), labels=seq(0,65,10))

ggsave("pdf_outputs/gby_pc_means.pdf")
```

Based on the data, PC1 x PC4 might be the best bet?
```{r}
pdf("pdf_outputs/carnatus_chrysomelas_pc1xpc4.pdf", width = 10, height = 10)
s.class(gby1_pca1$li, fac=pop(gby1), cell = 0,
        xax=1, yax=4, col=transp(c("dark blue", "dark green"),.8),
        axesel=FALSE, cstar=0, cpoint=4)
add.scatter.eig(gby1_pca1$eig[1:63],3,1,4, ratio=.3)
dev.off()

```

## Fst

```{r}
library(hierfstat)
```

```{r}
fst_gby <- pairwise.fst(gby1, pop = gby1$pop, res.type = c("matrix"))

fst_gby
```

This isn't properly accounting for the two species. Not sure why not.
```{r}
?basic.stats
gby_stats <- basic.stats(gby1, diploid = TRUE)

gby_per_loc <- rownames_to_column(gby_stats$perloc) %>%
  rename(locus = rowname)

gby_per_loc %>%
  arrange(desc(Fst)) %>%
  write_csv("csv_outputs/GBY_per_locus_stats.csv")

gby_stats$pop.freq$tag_id_179
  #write_csv("csv_outputs/gby_pop_freq.csv")
```



