---
title: "PCAs for manuscript"
output: html_notebook
---

Carlos requested that I make multiple PCAs with the clusters of subgenera for greater legibility, and that I keep the colors consistent throughout.

This also includes the Fst analysis.

I'll do that here.

```{r load-libraries}
library(tidyverse)
library(adegenet)
library("ape")
library("pegas")
library(RColorBrewer)

```

## Entire downsampled dataset

Load data for all 54 filtered species
```{r dataset-genepop}
dataset1 <- read.genepop("../new_baseline_data/processed/sebastes_final_dataset.gen", ncode = 3)

# set population as a factor
dataset1@pop <- as.factor(dataset1@pop)

```

### Test HWE

```{r}
# hw.test in pegas
# test_hw <- as.loci(dataset1)
# 
# practice <- by(test_hw, test_hw$population, hw.test)
# 
# (practice.mat <- sapply(practice, "[", i = TRUE, j = 4)) # Take the fourth column with all rows, which is Pr.exact
```

Test for HWE without permutations
```{r}
# dataset1.pop <- seppop(dataset1) %>% lapply(hw.test, B = 0)

```


```{r}
# dataset.mat <- sapply(dataset1.pop, "[", i = TRUE, j = 3) # Take the third column with all rows
# loci <- rownames(dataset.mat)
# p_vals <- as.tibble(dataset.mat)
# loci_names <- as.tibble(loci)
# hwe_spp <- bind_cols(loci_names, p_vals) %>%
#   rename(locus = value)
# 
# # I could tally how many loci deviate from HWE in each species
# 
# # save the table of HWE p-values for all 54 species
# hwe_spp %>%
#   write_csv("csv_outputs/HWE_spp_p_values.csv")
# 
# # how many loci out of hwe in each species/ which loci are out of hwe in which species
# hwe_spp %>%
#   gather(species, value = 2:55, -locus) %>%
#   rename(pval = `2:55`) %>%
#   filter(pval < 0.05) %>%
#   group_by(locus) %>%
#   tally() %>%
#   filter(n > 7)
```  

Which species are most affected by loci that are out of hwe?
```{r}
# hwe_spp %>%
#   gather(species, value = 2:55, -locus) %>%
#   rename(pval = `2:55`) %>%
#   filter(pval < 0.05) %>%
#   group_by(species) %>%
#   tally() %>%
#   arrange(desc(n))

```




```{r alpha-plot}
# alpha  <- 0.05 # set the alpha to 0.05
# newmat <- dataset.mat
# newmat[newmat > alpha] <- 1
# 
# # now create a simple heatmap to show where there are significant deviations from HWE
# library("lattice")
# levelplot(t(newmat))

```

## Checking pairwise Fst
```{r}
library(hierfstat)

fst_spp <- pairwise.fst(dataset1, pop = dataset1$pop, res.type = c("matrix"))

fst_summary <- as.tibble(fst_spp)
pairwise_fst <- rownames(fst_spp) %>%
  as.tibble() %>%
  rename(species = value) %>%
  bind_cols(fst_summary) 

# pairwise_fst %>%
#   write_csv("csv_outputs/sebastes_spp_pairwise_Fst.csv")

```

Make a distribution plot of Fst values between all pairs of species
```{r}
# reformat data
long_fst <- pairwise_fst %>%
  gather("pair", "Fst", 2:55, -species) %>%
  filter(Fst > 0) # get rid of the self-by-self comparisons

# there are a total of 1,425 unique comparisons (the symmetric matix has double that)

just_fst <- long_fst %>%
  arrange(desc(Fst)) %>%
  select(Fst) %>%
  unique()

# what's the mean and SD Fst values?
just_fst %>%
  summarise(mean(Fst))

# SD
just_fst %>%
  summarise(sd(Fst))

# plot the Fst values as a histogram
a <- just_fst %>%
  ggplot(aes(Fst))

a + geom_histogram(binwidth = 0.01) +
  geom_vline(aes(xintercept = mean(Fst)), 
             linetype = "dashed", size = 0.6) +
  theme_bw() +
  ylab("Count") +
  xlab("Pairwise Fst") 
  
# this figure includes double the number of actual comparisons
ggsave("pdf_outputs/pairwise_Fst_distribution.pdf", height = 4, width = 6)
```

Which pairwise comparisons are the highest Fst?
```{r}
long_fst %>%
  filter(Fst > 0.74) %>%
  group_by(Fst) %>%
  unique()

```

What are the comparisons among Pteropodus?
```{r}
long_fst %>%
  filter(species %in% c("atrovirens", "chrysomelas", "carnatus", "caurinus", "maliger", "nebulosus") & pair %in% c("chrysomelas", "carnatus", "caurinus", "maliger", "nebulosus")) %>%
  write_csv("csv_outputs/ptero_fst.csv")

v1 <- long_fst %>%
  filter(species == "atrovirens" & pair %in% c("chrysomelas", "carnatus", "caurinus", "maliger", "nebulosus"))

v2<- long_fst %>%
  filter(species == "chrysomelas" & pair %in% c("carnatus", "caurinus", "maliger", "nebulosus"))

v3 <- long_fst %>%
  filter(species == "carnatus" & pair %in% c("caurinus", "maliger", "nebulosus"))

v4 <- long_fst %>%
  filter(species == "caurinus" & pair %in% c("maliger", "nebulosus"))

v5 <- long_fst %>%
  filter(species == "maliger" & pair %in% c("nebulosus"))

bind_rows(v1, v2, v3, v4, v5) %>%
  arrange(desc(Fst)) %>%
  write_csv("csv_outputs/pteropodus_fst.csv")
```



```{r}
basic.stats(dataset1, diploid = TRUE)
```

## calculating linkage disequilibrium

```{r}

```




Make PCA
```{r dataset-pca}
# Allele presence absence data are extracted and NAs replaced using tab:
datasetX <- tab(dataset1, NA.method="mean")

# make PCA
dataset_pca1 <- dudi.pca(datasetX, center = TRUE, scannf = FALSE, scale=FALSE, nf = 1000)

# colors
mycol <- colorRampPalette(c("palegreen", "darkolivegreen", "darkslategrey", "darkgreen", "deepskyblue", "aquamarine4", "seagreen4", "cadetblue", "cyan", "darkblue", "midnightblue", "blue", "dodgerblue", "darkcyan", "darkslateblue", "slateblue", "steelblue4", "skyblue", "paleturquoise4", "brown", "royalblue", "purple4", "orange", "darkorange", "darkgoldenrod", "chocolate", "tan4", "saddlebrown", "sienna", "navajowhite4", "darkgray", "black"))(54)

# plot with factor labels
pdf("pdf_outputs/final_dataset_pca.pdf", width = 10, height = 10)
s.class(dataset_pca1$li, fac=pop(dataset1), wt = rep(1, length(pop(dataset1))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(mycol,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()

```

Amount of variance represented by each principal component
```{r}
eigen1 <- as.tibble(dataset_pca1$eig)

total_eigenval <- eigen1 %>%
  summarise(sum(value))

top32 <- eigen1 %>%
  top_n(., 32) %>%
  summarise(sum(value))
```


```{r}
top32/total_eigenval

```
77.5% of the total variance is captured in the first 32 eigenvalues.


```{r}
# PC1
1.712034e+01/total_eigenval
```
```{r}
# PC2
11.720844/total_eigenval
```

Using the adegenet manual to describe the PC variance in terms of percentage of the total variance:
```{r}
eig.perc <- 100*dataset_pca1$eig/sum(dataset_pca1$eig)

eig.perc
```
Great! Same answer. That validates the way I was thinking about the PCA.

Out of curosity, how much variation in the first 10 PCs?
```{r}
sum(eig.perc[1:10])
```
Beautiful! 50% of the variation is in the first 10 PCs.




I want to keep the colors for each species consistent, so I created an Excel document with that information `sebastes_spp_pca_colors.xlsx` in the sebastes_spp_ID folder.
```{r}
  as.tibble(mycol) %>% View
```


## Pteropodus

```{r pteropodus}
dataset2 <- read.genepop("../new_baseline_data/processed/pteropodus.gen", ncode = 3)

# set population as a factor
dataset2@pop <- as.factor(dataset2@pop)

# Allele presence absence data are extracted and NAs replaced using tab:
dataset2X <- tab(dataset2, NA.method="mean")

# make PCA
dataset2_pca <- dudi.pca(dataset2X, scannf = FALSE, center = TRUE, scale = FALSE, nf = 100)

# colors
mycol2 <- c("#98FB98", "#70A65A", "#4E6634", "#385547", "#1F5634", "#036205" ,"#009281", "#06BAF1", "#2E9BA0", "#3E8B6C", "#318B5B")

# plot with factor labels
pdf("pdf_outputs/pteropodus_pc1xpc2.pdf", width = 5, height = 5)
s.class(dataset2_pca$li, fac=pop(dataset2), wt = rep(1, length(pop(dataset2))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(mycol2,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()

```
The closest Pteropodus species are on top of each other: atrovirens, caurinus, chrysomelas, carnatus, maliger, etc.

PC 2 x PC 3 draws out atrovirens, but not the others
```{r}
s.class(dataset2_pca$li, fac=pop(dataset2), wt = rep(1, length(pop(dataset2))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=2, yax=3, col=transp(mycol,.8),
        axesel=FALSE, cstar=0, cpoint=1)
```

Let's quickly take a look to see if there are other PCs that do a better job?

## Mean of the values of each PC for Pteropodus
(based on my work in gopher/black-and-yellow)

```{r}
# turn df into a tibble,
# sort by the species groups, and then 

ptero_data <- rownames_to_column(dataset2_pca$li, var = "id") 
ptero_pop <- as.tibble(dataset2@pop) %>%
  rename(species = value)

# calculate the mean
pc_means <- bind_cols(ptero_pop, ptero_data) %>%
  group_by(species) %>%
  summarise_all(funs(mean)) %>%
  select(-id)

pc_means
```

So then I want to plot those means on the y-axis and have PC on the x-axis.

```{r}
mean_df <- pc_means %>%
  gather(pc, mean, 3:101)

# rename the Axis to just the PC number
mean_df$pc <- str_replace(mean_df$pc, "Axis", "")
mean_df$pc <- as.numeric(mean_df$pc)
```

```{r means-plot}
mean_df %>%
  ggplot() +
  geom_point(aes(x = pc, y = mean, color = species)) +
  theme_bw()

#ggsave("pdf_outputs/pteropodus_pc_means.pdf")
```

Try making a dataset with just the tightest cluster from Pteropodus:

# Pteropodus subset

```{r}
ptero_sub <- read.genepop("../new_baseline_data/processed/pteropodus_subset.gen", ncode = 3)

# set population as a factor
ptero_sub@pop <- as.factor(ptero_sub@pop)

# Allele presence absence data are extracted and NAs replaced using tab:
ptero_subX <- tab(ptero_sub, NA.method="mean")

# make PCA
ptero_sub_pca <- dudi.pca(ptero_subX, scannf = FALSE, center = TRUE, scale = FALSE, nf = 10)

# colors
ptero_mycol <- c("#98FB98", "#70A65A", "#4E6634", "#385547", "#1F5634", "#009281", "#06BAF1", "#2E9BA0")

# plot with factor labels
pdf("pdf_outputs/ptero_sub_pca.pdf", width = 5, height = 5)
s.class(ptero_sub_pca$li, fac=pop(ptero_sub), wt = rep(1, length(pop(ptero_sub))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(ptero_mycol,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()
```

Pteropodus subset v2
```{r}
ptero_sub2 <- read.genepop("../new_baseline_data/processed/pteropodus_subset_v2.gen", ncode = 3)

# set population as a factor
ptero_sub2@pop <- as.factor(ptero_sub2@pop)

# Allele presence absence data are extracted and NAs replaced using tab:
ptero_sub2X <- tab(ptero_sub2, NA.method="mean")

# make PCA
ptero_sub2_pca <- dudi.pca(ptero_sub2X, scannf = FALSE, center = TRUE, scale = FALSE, nf = 10)

# colors
ptero2_mycol <- c("#98FB98", "#70A65A", "#4E6634", "#385547", "#1F5634")

# plot with factor labels
pdf("pdf_outputs/ptero_subset_v2.pdf", width = 5, height = 5)
s.class(ptero_sub2_pca$li, fac=pop(ptero_sub2), wt = rep(1, length(pop(ptero_sub2))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(ptero2_mycol,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()

```



## Sebastomus

read in a genepop file with just those species overlapping in the middle of the plot
```{r}
dataset3 <- read.genepop("../new_baseline_data/processed/sebastomus.gen", ncode = 3)

# set population as a factor
dataset3@pop <- as.factor(dataset3@pop)

# Allele presence absence data are extracted and NAs replaced using tab:
dataset3X <- tab(dataset3, NA.method="mean")

# make PCA
dataset3_pca <- dudi.pca(dataset3X, scannf = FALSE, center = TRUE, scale = FALSE, nf = 10)

# colors
sebtocol <- c("#439376", "#5D9FA1", "#25D8D9", "#00CEE9", "#0039A5", "#080881", "#0B0BBB", "#5F5CBF", "#406198", "#548CAF", "#FF9000", "#8B4615","#954B1F", "#9E5430", "#0310FF", "#8B5323")

# plot with factor labels
pdf("pdf_outputs/sebastomus_pca.pdf", width = 8, height = 8)
s.class(dataset3_pca$li, fac=pop(dataset3), wt = rep(1, length(pop(dataset3))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(sebtocol,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()

```
To get better resolution, I'll look at two separate clusters in this group.

## Sebastomus cluster 1

This is a subset of the above group that needs additional separation.

```{r sebastomus-cluster1-genepop}
sebastomus1 <- read.genepop("../new_baseline_data/processed/sebastomus_cluster1.gen", ncode = 3)

# set population as a factor
sebastomus1@pop <- as.factor(sebastomus1@pop)

# Allele presence absence data are extracted and NAs replaced using tab:
sebastomus1X <- tab(sebastomus1, NA.method="mean")

# make PCA
sebastomus1_pca <- dudi.pca(sebastomus1X, scannf = FALSE, center = TRUE, scale = FALSE, nf = 10)

# colors
seb1col <- c("#00CEE9", "#0039A5", "#080881", "#171771", "#0B0BBB", "#0310FF", "#5F5CBF")

# plot with factor labels
pdf("pdf_outputs/sebastomus_cluster1_pc2x3.pdf", width = 5, height = 5)
s.class(sebastomus1_pca$li, fac=pop(sebastomus1), wt = rep(1, length(pop(sebastomus1))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=2, yax=3, col=transp(seb1col,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()
```

## Sebastomus cluster 2
This one includes:
aleutianus, melanostictus, ruberrimus, polyspinis, and pinniger
```{r sebastomus-cluster2}
sebastomus2 <- read.genepop("../new_baseline_data/processed/sebastomus_cluster2.gen", ncode = 3)

# set population as a factor
sebastomus2@pop <- as.factor(sebastomus2@pop)

# Allele presence absence data are extracted and NAs replaced using tab:
sebastomus2X <- tab(sebastomus2, NA.method="mean")

# make PCA
sebastomus2_pca <- dudi.pca(sebastomus2X, scannf = FALSE, center = TRUE, scale = FALSE, nf = 10)

# colors
seb_c2col <- c("#25D8D9", "#FF9000", "#954B1F", "#9E5430", "#8B5323")

# plot with factor labels
pdf("pdf_outputs/sebastomus_cluster2_pca.pdf", width = 5, height = 5)
s.class(sebastomus2_pca$li, fac=pop(sebastomus2), wt = rep(1, length(pop(sebastomus2))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(seb_c2col,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()
```


## Sebastosomus

read in a genepop file
```{r}
dataset4 <- read.genepop("../new_baseline_data/processed/sebastosomus.gen", ncode = 3)

# set population as a factor
dataset4@pop <- as.factor(dataset4@pop)

# Allele presence absence data are extracted and NAs replaced using tab:
dataset4X <- tab(dataset4, NA.method="mean")

# make PCA
dataset4_pca <- dudi.pca(dataset4X, scannf = FALSE, center = TRUE, scale = FALSE, nf = 10)

# colors
mycol <- c("#863D61", "#4C61CC", "#4A43B8", "#5E2183", "#C27331", "#FF9F00")

# plot with factor labels
pdf("pdf_outputs/sebastosomus_pca.pdf", width = 8, height = 8)
s.class(dataset4_pca$li, fac=pop(dataset4), wt = rep(1, length(pop(dataset4))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(mycol,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()

```




## Sebastodes

Maybe having smaller groups of species will help?
```{r sebastodes-genepop}
dataset5 <- read.genepop("../new_baseline_data/processed/sebastodes.gen", ncode = 3)

# set population as a factor
dataset5@pop <- as.factor(dataset5@pop)

# Allele presence absence data are extracted and NAs replaced using tab:
dataset5X <- tab(dataset5, NA.method="mean")

# make PCA
dataset5_pca <- dudi.pca(dataset5X, scannf = FALSE, center = TRUE, scale = FALSE, nf = 10)

# colors
seb2col <- c("#83CAE7", "#6E7E7E", "#934545", "#E28904")

# plot with factor labels
pdf("pdf_outputs/sebastodes_pca.pdf", width = 5, height = 5)
s.class(dataset5_pca$li, fac=pop(dataset5), wt = rep(1, length(pop(dataset5))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(seb2col,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()
```


## Eosebastes

species in this PCA include:
melanostomus
reedi
crameri
rufus
ovalis
```{r eosebastes-genepop}
eosebastes <- read.genepop("../new_baseline_data/processed/eosebastes.gen", ncode = 3)

# set population as a factor
eosebastes@pop <- as.factor(eosebastes@pop)

# Allele presence absence data are extracted and NAs replaced using tab:
eosebastesX <- tab(eosebastes, NA.method="mean")

# make PCA
eosebastes_pca <- dudi.pca(eosebastesX, scannf = FALSE, center = TRUE, scale = FALSE, nf = 10)

# colors
eo_col <- c("#493E8D", "#5D4FB4", "#C67515", "#C7661F", "#9D5D27")

# plot with factor labels
pdf("pdf_outputs/eosebastes_pca.pdf", width = 5, height = 5)
s.class(eosebastes_pca$li, fac=pop(eosebastes), wt = rep(1, length(pop(eosebastes))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(eo_col,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()
```


## Can I filter the original genind object or data?

```{r}
dataset_pcs <- as.tibble(dataset_pca1$li)
dataset_pops <- as.tibble(dataset1@pop) %>%
  rename(species = value)

full_tbl <- bind_cols(dataset_pops, dataset_pcs)
```


```{r}
full_tbl %>%
  filter(species %in% c("atrovirens", "carnatus", "chrysomelas", "maliger", "nebulosus")) %>%
  ggplot(aes(x = Axis1, y = Axis2, color = species)) +
  geom_point()
  
```
That doesn't look good.


## Mismash

This group includes species from multiple subgenera...
```{r}
dataset6 <- read.genepop("../new_baseline_data/processed/mismash.gen", ncode = 3)

# set population as a factor
dataset6@pop <- as.factor(dataset6@pop)

# Allele presence absence data are extracted and NAs replaced using tab:
dataset6X <- tab(dataset6, NA.method="mean")

# make PCA
dataset6_pca <- dudi.pca(dataset6X, scannf = FALSE, center = TRUE, scale = FALSE, nf = 10)

# colors
mashcol <- c("#1464FF","#158EDE", "#038B9A", "#20678B", "#B9860A")

# plot with factor labels
pdf("pdf_outputs/mismash_pca.pdf", width = 5, height = 5)
s.class(dataset6_pca$li, fac=pop(dataset6), wt = rep(1, length(pop(dataset6))), clabel = .8, grid = FALSE, cellipse = 2,
        xax=1, yax=2, col=transp(mashcol,.8),
        axesel=FALSE, cstar=0, cpoint=1)
dev.off()
```


